\paperTitle

### 2014-01-31 doc/icml2014/abstract.tex
Updated figures and some details and notation
Percy's changes

\begin{abstract} 
\end{abstract}

\paragraph{Errata}
\begin{enumerate}
  \item The likelihoods as defined are concave, not convex.
  \item Define parameters $\theta$ more uniformly.
\end{enumerate}

>>>>

\begin{abstract}

Recent work in method of moments provide consistent estimates
  for latent-variable models, avoiding local optima issues,
  but these methods can only be applied to certain types of graphical models.
In this work, we show that the method of moments in conjunction
  with a composite marginal likelihood objective
  yields consistent parameter estimates for a much broader
  class of directed and undirected graphical models,
  including loopy graphs with high treewidth.
%In particular, our algorithm applies to any graphical model wherein each
  %hidden variable has at least three conditionally independent observed
  %variables.
  %that sidestep local optima issues for a general class of
  %directed and undirected graphical models.
%Parameter estimation in latent-variable models based on EM
  %is prone to local optima.
  %due to the non-convexity of the marginal log-likelihood. 
%In this paper, we extend recent advances in the method of moments 
  %to provide consistent parameter estimates
  %that sidestep local optima issues for a general class of
  %directed and undirected graphical models.
%In particular, our algorithm applies to any graphical model wherein each
  %hidden variable has at least three conditionally independent observed
  %variables.
%This family includes loopy graphs with high treewidth.
Specifically, we use tensor factorization to reveal partial information
about the latent variables, rendering the otherwise non-convex negative
log-likelihood convex.
%We use the method of moments to tie each hidden to some observed
  %variables, and show that the rest of the problem can be cast as convex
  %optimization of the likelihood.
Our approach gracefully extends to models outside our class by
  incorporating the partial information via posterior regulraization.
\end{abstract}

\paragraph{Last updated:} \today, \currenttime

\paragraph{Errata / Incomplete}
\begin{enumerate}
  \item Use \texttt{\textbackslash left(}.
  \item Use $\ell$ instead of $L$.
  \item convex and negative log likelihood.
\end{enumerate}

### 2014-01-31 doc/icml2014/background.tex
Percy's changes

Let $\sG$ be a discrete graphical model with observed variables $X$ and
  hidden variables $H$.
Let $\Pa(a) \subset X \union H$ be the parents of a variable $a$ in a directed
  model, and $\sN(a) \subset X \union H$ be the neighbours of $a$ in an
  undirected model.
We will use $x_1, x_2, \cdots, x_L \in X$ to denote observed variables and
  $h_1, h_2, \cdots, h_M \in H$ to denote hidden variables.
For ease of exposition, we assume that $x_i \in \{1, \cdots, d\}$ for
  all $i$, and that $h_i \in \{1, \cdots, k\}$ for all $i$.
Let $\sX$ be the joint domain of the observed variables, i.e. $\sX \eqdef
  \underbrace{d \times d \cdots \times d}_{L ~\textrm{times}}$ and
let $\sH$ be the joint domain of the hidden variables, i.e. $\sH \eqdef
  \underbrace{k \times k \cdots \times k}_{M ~\textrm{times}}$.

We assume that the model $\sG$ is parameterized at every clique $\sC$.
If $\sG$ is directed, the parameters correspond to
  the conditional probability tables for each clique.
If $\sG$ is undirected, the parameters correspond to the clique
  potentials.

@@

% Maximum likelihood
The standard estimation procedure is maximum (marginal) likelihood,
  \begin{align*}
    \sL &\eqdef \max_{\theta} \sum_{\vx \in \sD} \log p_\theta(\vx) \\
        &=      \max_{\theta} \sum_{\vx \in \sD} \log \sum_{\vh \in \sH} p_\theta(\vx,\vh)
  \end{align*}
  which is statistically efficient but computationally intractable
  because we must marginalize over latent variables $\vh$.
In practice, one uses gradient-based optimization procedures (e.g., EM
  or L-BFGS) on the marginal likelihood, which can get stuck in local
  optima.

@@

We use the notation $[\cdot]$ to indicate indexing; for example $M[i]$
  is the $i$-th row of a matrix $M$ and $M[i,j]$ is the $i,j$-th element
  of $M$.
For a tensor $T \in \Re^{d \times \ldots \times d}$ and a vector $\vec
  i = (i_1, \cdots, i_l)$, $T[\vec i] = T_{i_1 \ldots i_l}$.
Similarly, for a set $\sC$ of size $k$ and vector $\vec x \in \Re^d$, we
  will use the notation $\vec x_\sC$ to get a vector $\Re^k$ which selects
  the appropriate elements of $\vec x$

We use $\otimes$ to denote the tensor product, e.g if $u \in \Re^k,
  v \in \Re^d$, then $u \otimes v \in \Re^{k \times d}$.
For a $\ell$-th order tensor $T \in \Re^{d \times \ldots \times
  d}$ and vectors $x_1, \cdots, x_\ell \in \Re^{d},$, we define 
\begin{align*}
  T(x_1, \cdots, x_\ell) 
      &= \sum_{i_1 = 1}^{d} \cdots \sum_{i_\ell = 1}^{d} 
            T[i_1, i_2, \ldots, i_l] x_1[i_1] \cdots x_\ell[i_\ell].
\end{align*}
Analogously, for a matrices $M_1 \in \Re^{d \times k}, \cdots,
  M_\ell \in \Re^{d \times k}$, we define
\begin{align*}
  T(M_1, \cdots, M_\ell)[\vec j]
      &= \sum_{i_1 = 1}^{d} \cdots \sum_{i_\ell = 1}^{d} 
            T[\vec i] {M_1}[i_1,j_1] \cdots M_\ell[i_\ell, j_\ell],
\end{align*}
where $\vec i = (i_1, \cdots, i_\ell)$ and $\vec j = (i_1, \cdots, j_\ell)$

To denote the moments of observed variables, we will use the notation
\begin{align*}
  \mO_i &\eqdef \Pr(x_i), &
  \mO_{ij} &\eqdef \Pr(x_i, x_j), &
  \mO_{ijk} &\eqdef \Pr(x_i, x_j, x_k),
\end{align*}
and so on.
We will view $\mO_i, \mO_{ij}, \mO_{ijk}, \cdots$ as tensors of
  order 1, 2, 3, \ldots in $\Re^d, \Re^{d\times d}, \Re^{d \times
  d \times d},\ldots$.
For the moments of hidden variables, we will use 
\begin{align*}
  \mH_i &\eqdef \Pr(h_i), &
  \mH_{ij} &\eqdef \Pr(h_i, h_j), &
  \mH_{ijk} &\eqdef \Pr(h_i, h_j, h_k).
\end{align*}

@@

Recently, \citet{anandkumar12moments} presented a consistent algorithm
  using the method of moments to learn the parameters of a three-view
  mixture model (\figureref{three-view}), an important case of directed
  graphical models, that sidesteps the local optima problem altogether. 
In the model, we have three observed variables $x_1, x_2, x_3$
  that are independent conditioned on a latent variable $h$.  
The parameters of this model are the mixture probabilities $\pi\oft 1 \eqdef
  \Pr(h) \in \Delta_{k-1}$ and the conditional moments $\mOpp{i}{1} \eqdef
  \Pr(x_i \given h) \in \Re^{d \times k}$ for $i \in \{1,2,3\}$.

  The key idea is that the observed moments $\mO_{12}, \mO_{23}, \mO_{13}$ and
  $\mO_{123}$ admit a factorization in terms of parameters;
\begin{align*}
  \mO_{ij} &= \sum_h {\pi \oft 1}[h] {\mOppt{i}{1}}[h] \otimes {\mOppt{j}{1}}[h] \quad \forall i,j \in \{1,2,3\} \\
  \mO_{123} &= \sum_h {\pi \oft 1}[h] {\mOppt{i}{1}}[h] \otimes {\mOppt{2}{1}}[h] \otimes  {\mOppt{2}{1}}[h].
\end{align*}

\citet{anandkumar13tensor} presented a robust tensor power method, which
  we call $\TensorFactorize$, which converts consistent estimates of the observed moments 
  into consistent estimates of the parameters, $\pi\oft 1, \mOpp{1}{1},
  \mOpp{2}{1},\mOpp{3}{1}$ (up to permutation of the columns), provided
  the following assumption holds:

@@

In brief, the algorithm first computes a whitening matrix $W \in
  \Re^{d \times k}$ such that $W^T \mO_{12} W = I_{k \times k}$,
  and uses it to transform $\mO_{123}$ into a symmetric orthogonal tensor.  
Then a robust tensor power method is applied to extract the eigenvectors
of the whitened $\mO_{123}$; unwhitening yields the columns of $\mOpp{3}{1}$.
  $\mOpp{1}{1}$ and $\mOpp{2}{1}$ can then be recovered with some algebra.

>>>>

Let $\sG$ be a discrete graphical model with observed variables $X$ and
  hidden variables $H$.
Let $\Pa(a) \subset X \union H$ be the parents of a variable $a$ in a directed
  model, and $\sN(a) \subset X \union H$ be the neighbors of $a$ in an
  undirected model.
We will use $x_1, x_2, \cdots, x_L \in X$ to denote observed variables and
  $h_1, h_2, \cdots, h_M \in H$ to denote hidden variables.
For an integer $n$, let $[n] = \{ 1, \dots, n \}$.
For ease of exposition, we assume that $x_i \in [d]$ for
all $i$, and that $h_i \in [k]$ for all $i$.
Let $\sX$ be the joint domain of the observed variables, i.e. $\sX \eqdef [d]^L$ and
let $\sH$ be the joint domain of the hidden variables, i.e. $\sH \eqdef [k]^M$.

% PL: don't know what this means
%We assume that the model $\sG$ is parameterized at every clique $\sC$.
If $\sG$ is directed, the model parameters correspond to
the conditional probability tables for each clique $(\Pa(a), a) \in \sG$.
If $\sG$ is undirected, we assume a log-linear parametrization of each clique $\sC \in \sG$.

@@

% Maximum likelihood
The standard estimation procedure is maximum (marginal) likelihood,
  \begin{align*}
    \sL &\eqdef \max_{\theta} \sum_{\vx \in \sD} \log p_\theta(\vx) \\
        &=      \max_{\theta} \sum_{\vx \in \sD} \log \sum_{\vh \in \sH} p_\theta(\vx,\vh).
  \end{align*}
  Maximum likelihood is statistically efficient,
  but in general computationally intractable
  because marginalizing over latent variables $\vh$ yields a non-convex objective.
In practice, one uses local optimization procedures (e.g., EM
  or L-BFGS) on the marginal likelihood, which can get stuck in local
  optima.

@@

We use the notation $[\cdot]$ to indicate indexing; for example $M[i]$
  is the $i$-th row of a matrix $M$ and $M[i,j]$ is the $i,j$-th element
  of $M$.
For a tensor $T \in \Re^{d \times \ldots \times d}$ and a vector $\vec
  i = (i_1, \cdots, i_l)$, define the projection $T[\vec i] = T_{i_1 \ldots i_l}$.
Similarly, for a set $\sC$ of size $k$ and vector $\vec x \in \Re^d$, we
  will use the notation $\vec x_\sC$ to get a vector $\Re^k$ which selects
  the appropriate elements of $\vec x$.

We use $\otimes$ to denote the tensor product: if $u \in \Re^d,
  v \in \Re^k$, then $u \otimes v \in \Re^{d \times k}$.
For an $\ell$-th order tensor $T \in \Re^{d \times \ldots \times
  d}$ and vectors $x_1, \cdots, x_\ell \in \Re^{d}$, define 
  the partial application:
\begin{align*}
  T(x_1, \cdots, x_\ell) 
      %&= \sum_{i_1 = 1}^{d} \cdots \sum_{i_\ell = 1}^{d} 
  &= \sum_{\bi}
            T[i_1, i_2, \ldots, i_l] x_1[i_1] \cdots x_\ell[i_\ell].
\end{align*}
Analogously, for a matrices $M_1 \in \Re^{d \times k}, \cdots,
  M_\ell \in \Re^{d \times k}$, we define
\begin{align*}
  T(M_1, \cdots, M_\ell)[\vec j]
      %&= \sum_{i_1 = 1}^{d} \cdots \sum_{i_\ell = 1}^{d} 
  &= \sum_{\bi}
            T[\vec i] {M_1}[i_1,j_1] \cdots M_\ell[i_\ell, j_\ell].
\end{align*}
%where $\vec i = (i_1, \cdots, i_\ell)$ and $\vec j = (i_1, \cdots, j_\ell)$

To denote the moments, we will use the notation:
\begin{align*}
  \mO_i &\eqdef \Pr(x_i), &
  \mO_{ij} &\eqdef \Pr(x_i, x_j), &
  \mO_{ijk} &\eqdef \Pr(x_i, x_j, x_k),
\end{align*}
and so on.
We will view $\mO_i, \mO_{ij}, \mO_{ijk}, \cdots$ as tensors of
  order 1, 2, 3, \ldots in $\Re^d, \Re^{d\times d}, \Re^{d \times
  d \times d},\ldots$.
For the latent marginals, we will use 
\begin{align*}
  \mH_i &\eqdef \Pr(h_i), &
  \mH_{ij} &\eqdef \Pr(h_i, h_j), &
  \mH_{ijk} &\eqdef \Pr(h_i, h_j, h_k).
\end{align*}

@@

\citet{anandkumar12moments} presented a consistent estimator
  using the method of moments to learn the parameters of a three-view
  mixture model (\figureref{three-view}), an important case of directed
  graphical models. %, that sidesteps the local optima problem altogether. 
In that model, we have three observed variables $x_1, x_2, x_3$
  that are independent conditioned on a latent variable $h_1$.  
The parameters of this model are the mixture probabilities $\pi\oft 1 \eqdef
  \Pr(h) \in \Delta_{k-1}$ and the conditional moments $\mOpp{i}{1} \eqdef
  \Pr(x_i \given h) \in \Re^{d \times k}$ for $i \in \{1,2,3\}$.

  The key idea is that the observed moments $\mO_{12}, \mO_{23}, \mO_{13}$ and
  $\mO_{123}$ admit a factorization in terms of parameters:
\begin{align*}
  \mO_{ij} &= \sum_h {\pi \oft 1}[h] {\mOpp{i}{1}}[h] \otimes {\mOpp{j}{1}}[h] \quad \forall i,j \in \{1,2,3\} \\
  \mO_{123} &= \sum_h {\pi \oft 1}[h] {\mOpp{1}{1}}[h] \otimes {\mOpp{2}{1}}[h] \otimes  {\mOpp{3}{1}}[h].
\end{align*}

\citet{anandkumar13tensor} presented a robust tensor power method, which
  we call $\TensorFactorize$, which converts estimates of the observed moments 
  into estimates of the parameters, $\pi\oft 1, \mOpp{1}{1},
  \mOpp{2}{1},\mOpp{3}{1}$ (up to permutation of the columns), provided
  the following assumption holds:

@@

In brief, the algorithm first computes a whitening matrix $W \in
  \Re^{d \times k}$ such that $W^T \mO_{12} W = I_{k \times k}$,
  and uses it to transform $\mO_{123}$ into a symmetric orthogonal tensor.  
Then a robust tensor power method is applied to extract the eigenvectors
of the whitened $\mO_{123}$; unwhitening yields the columns of $\mOpp{3}{1}$.
  $\mOpp{1}{1}$ and $\mOpp{2}{1}$ can then be recovered with a few matrix operations.

The tensor factorization method attacks the heart of the non-convexity
  in latent-variable models.  In the next two sections,
  we will see how to build on top of this result to estimate directed (Section~\ref{sec:directed})
  and undirected (Section~\ref{sec:undirected}) models.
